### 2013-06-27 drafts/planner/planner_inference.tex
Updated writeup on planner

\section{Inference}
Following the Monte Carlo EM derivation described in the ``inference'' document, we want to find the grammar maximizing
$$
\log P(G|\theta)+\sum_n \mathbb{E}_{q_n}\left[ \log P(\rho_n | G) \right]
$$
where
$$
\tilde{q}_n(e_1\cdots e_k) = P(t_n | e_1\cdots e_k)\prod_i P(e_i | G)
$$
Now, if we can sample from $\tilde{q}$, we're done.
We could just sample/enumerate random plans from $G$ and then do likelihood weighting, but if our frontier is of order $10^4$ or so, this will quickly become impractical for $\ell$ of order $10^1$.
Instead, we'll use MCMC to get samples from a proposal distribution, which we'll then reweight.
\subsection{Planning via MCMC}
We'll draw samples (plans) from a distribution $\tilde{g}(\cdot)$ that will serve as a proposal distribution for the expectation w.r.t. $q_n$.
This MCMC algorithm is not Metropolis-Hastings, but it has a stationary distribution which is similar to $q_n$.
Given a plan $\rho$ with $|\rho|<\ell$, the proposal distribution $h(\cdot | \cdot)$ is
$$
h(\rho e | \rho) \propto P(t|\rho e) P(e|G)
$$
A plan $|\rho|=\ell$ has $h(\emptyset | \rho)=1$.
Intuitively, the sampler is growing out a plan, starting from $\emptyset$, sampling the next move according to the posterior distribution upon the next program (greedy one-step lookahead).
The stationary distribution of this Markov chain is
$$
\tilde{g} (e_1 e_2 \cdots e_k) = P(e_1 | G) P(e_2 | G) \cdots P(e_k | G) P(t | e_1) P(t | e_1e_2) \cdots P(t | e_1e_2\cdots e_k)
$$
as can be verified by observing that $\tilde{g}(\rho e)=h(\rho e | \rho) \tilde{g}(\rho)$.
So, we reweight a plan $\rho=e_1e_2\cdots e_k$ by the quantity
$$
\frac{\tilde{q}}{\tilde{g}} = \frac{1}{P(t|e_1)P(t|e_1e_2)\cdots P(t|e_1e_2\cdots e_{k-1})}
$$
Each plan of length $\ell$ sampled contributes $\ell$ total samples (all prefixes of the plan).
Each total plan $\rho=e_1\cdots e_\ell$ contributes the following quantity to the maximization problem in $G$ (up to a normalizing constant):
\begin{eqnarray*}
&&\sum_{k\leq\ell} \frac{\tilde{q}(e_1\cdots e_k)}{\tilde{g}(e_1\cdots e_k)} \sum_{j\leq k} \log P(e_j | G)\\
&=&\sum_{k\leq\ell} \frac{1}{P(t|e_1)P(t|e_1e_2)\cdots P(t|e_1e_2\cdots e_{k-1})} \sum_{j\leq k} \log P(e_j | G)\\
&=&\sum_{j\leq \ell} w_j\log P(e_j | G)
\end{eqnarray*}
where I've introduced the quantities
$$
w_j=\sum_{m\geq j} \frac{1}{P(t|e_1)P(t|e_1e_2)\cdots P(t|e_1\cdots e_{j-1})}
$$
The $w_j$'s factor in a nice way that allows them to be computed in linear time using a dynamic program.
\footnote{The sums for $w_j$ end up being the partition function if $j=1$ and is a sort of ``truncated partition function'' for $j>1$.
For the $j>1$ case, the weights $w_j$ are what the partition function \emph{would have been} if the plan had started with programs $e_{k<j}$ for free.}
The following procedure samples a plan and returns the weights upon each program in that plan, along with the partition function:
\begin{codebox}
\Procname{$\proc{Sample-Plan}(\rho)$}
\li \Comment{Draw the next program in the plan}
\li \id{e} $\sim$ $P(\rho\cdot | t, G)$
\li \Comment{If the plan should be extended, recurse, creating the suffix of the plan}
\li \If $|\rho e|<\ell$ \Then
\li \id{suffixWeight}, \id{suffixRewards} = \func{Sample-Plan}$(\rho e)$
\li \Else
\li \id{suffixWeight}, \id{suffixRewards} = $0$, $\emptyset$
\li \End
\li \Comment{To satisfy the boundary conditions of the recurrence, $P(t|\emptyset)$ is defined to be 1.}
\li \id{eReward} = $\frac{1+\mbox{\id{suffixRewards}}}{P(t|\rho)}$
\li \Return \id{eReward}, $\{(e, \mbox{\id{eReward}})\}\cup \mbox{\id{suffixRewards}}$
\end{codebox}

@@

The likelihood isn't the only heuristic that could be used in the stochastic greedy search.
If we had some other heuristic, $\hat{\mbox{H}}$, one reasonable proposal for the MCMC walk would be the Boltzman distribution:
$$
h(\rho e | \rho) \propto \exp\left(-\hat{\mbox{H}}(\rho e)\right + \log P(e|G))
$$

>>>>

\section{Inference}
Following the Monte Carlo EM derivation described in the ``inference'' document, we want to find the grammar maximizing
$$
\log P(G|\theta)+\sum_n \mathbb{E}_{q_n}\left[ \log P(\rho_n | G) \right]
$$
where
$$
\tilde{q}_n(e_1\cdots e_k) = P(t_n | e_1\cdots e_k)\prod_i P(e_i | G)
$$
Now, if we can sample from $\tilde{q}$, we're done.
We could just sample/enumerate random plans from $G$ and then do likelihood weighting, but if our frontier is of order $10^4$ or so, this will quickly become impractical for $\ell$ of order $10^1$.
Instead, we'll use MCMC to get samples from a proposal distribution, which we'll then reweight.
\subsection{Planning via MCMC}
We'll draw samples (plans) from a distribution $\tilde{g}(\cdot)$ that will serve as a proposal distribution for the expectation w.r.t. $q_n$.
This MCMC algorithm is not Metropolis-Hastings, but it has a stationary distribution which is similar to $q_n$.
Given a plan $\rho$ with $|\rho|<\ell$, the proposal distribution $h(\cdot | \cdot)$ is
$$
h(\rho e | \rho) \propto P(t|\rho e) P(e|G)
$$
A plan $|\rho|=\ell$ has $h(\emptyset | \rho)=1$.
Intuitively, the sampler is growing out a plan, starting from $\emptyset$, sampling the next move according to the posterior distribution upon the next program (greedy one-step lookahead).
The stationary distribution of this Markov chain is
$$
\tilde{g} (e_1 e_2 \cdots e_k) = P(e_1 | G) P(e_2 | G) \cdots P(e_k | G) P(t | e_1) P(t | e_1e_2) \cdots P(t | e_1e_2\cdots e_k)
$$
as can be verified by observing that $\tilde{g}(\rho e)=h(\rho e | \rho) \tilde{g}(\rho)$.
So, we reweight a plan $\rho=e_1e_2\cdots e_k$ by the quantity
$$
\frac{\tilde{q}}{\tilde{g}} = \frac{1}{P(t|e_1)P(t|e_1e_2)\cdots P(t|e_1e_2\cdots e_{k-1})}
$$
Each plan of length $\ell$ sampled contributes $\ell$ total samples (all prefixes of the plan).
Each total plan $\rho=e_1\cdots e_\ell$ contributes the following quantity to the maximization problem in $G$ (up to a normalizing constant):
\begin{eqnarray*}
&&\sum_{k\leq\ell} \frac{\tilde{q}(e_1\cdots e_k)}{\tilde{g}(e_1\cdots e_k)} \sum_{j\leq k} \log P(e_j | G)\\
&=&\sum_{k\leq\ell} \frac{1}{P(t|e_1)P(t|e_1e_2)\cdots P(t|e_1e_2\cdots e_{k-1})} \sum_{j\leq k} \log P(e_j | G)\\
&=&\sum_{j\leq \ell} w_j\log P(e_j | G)
\end{eqnarray*}
where I've introduced the quantities
$$
w_j=\sum_{m\geq j} \frac{1}{P(t|e_1)P(t|e_1e_2)\cdots P(t|e_1\cdots e_{j-1})}
$$
The $w_j$'s factor in a nice way that allows them to be computed in linear time using a dynamic program.
\footnote{The sums for $w_j$ end up being the partition function if $j=1$ and is a sort of ``truncated partition function'' for $j>1$.
For the $j>1$ case, the weights $w_j$ are what the partition function \emph{would have been} if the plan had started with programs $e_{k<j}$ for free.}
The following procedure samples a plan and returns the weights upon each program in that plan, along with the partition function:
\begin{codebox}
\Procname{$\proc{MCMC-Plan}(\rho, \mbox{\id{reciprocalLikelihoods}})$}
\li \Comment{Draw the next program in the plan}
\li \id{e} $\sim$ $P(\rho\cdot | t, G)$
\li \Comment{If the plan should be extended, recurse, creating the suffix of the plan}
\li \If $|\rho e|<\ell$ \Then
\li \id{suffixZ}, \id{suffixWeights} = \func{MCMC-Plan}$(\rho e, \mbox{\id{reciprocalLikelihoods}}/P(t|\rho e))$
\li \Else
\li \id{suffixZ}, \id{suffixWeights} = $0$, $\emptyset$
\li \End
\li \id{Z} = \id{suffixZ} + \id{reciprocalLikelihoods}
\li \Return \id{Z}, $\{(e, \mbox{\id{Z}})\}\cup \mbox{\id{suffixWeights}}$
\end{codebox}
where \emph{reciprocalLikelihoods} is set to $1$ initially.

@@

The likelihood isn't the only heuristic that could be used in the stochastic greedy search.
If we had some other heuristic, $\hat{\mbox{H}}$, one reasonable proposal for the MCMC walk would be the Boltzman distribution:
$$
h(\rho e | \rho) \propto \exp\left(-\hat{\mbox{H}}(\rho e) + \log P(e|G)\right)
$$
